### 2017-02-25 2017mergingmerges/section/XX-Related.tex
related work

Existing systems for combinator fusion only support a subset of combinators, and only in limited cases.
Pull fusion supports joins where one combinator pulls from multiple inputs, but not splits where one output is used my multiple combinators.
Pull can thus express @merge@, but not @partition@.
Push fusion supports splits (@partition@), but not joins (@merge@).
Polarised fusion~\cite{lippmeier2016polarized} addresses these shortcomings by separating the computation into pull producers and push consumers, with explicit drain loops to convert between the two sections. However this requires manual plumbing.

The duality between pull and push arrays has also been explored in Obsidian by \citep{claessen2012expressive} and later in \citep{svensson2014defunctionalizing}, by explicitly separating the types of pull and push arrays.
Here the distinction is made for the purpose of code generation for GPUs rather than fusion, as operations like append on pull arrays require conditionals inside the loop, whereas using push arrays moves these conditionals outside the loop.
% Here push arrays are `defunctionalized' and represented as a datatype containing the combinator and its arguments, which allows specialised and efficient code generation.
% However this work only supports a limited number of baked-in push combinators such as append, mapping, interleaving and reversing.

Streaming IO libraries have blossomed in the Haskell ecosystem, generally based on Iteratees \cite{kiselyov2012iteratees}.
Libraries such as @conduit@ \cite{hackage:conduit}, @enumerator@ \cite{hackage:enumerator}, @machines@ \cite{hackage:machines} and @pipes@ \cite{hackage:pipes} are all designed to write stream computations with bounded buffers.
However, these libraries are not designed for absolute performance and programs tend to be written to operate over streams of chunks rather than streams of values.
For the most part they support only simple straight-line computations.

Another way to characterise fusion systems is whether they are \emph{local} (shortcut) or \emph{global} fusion.
Local fusion systems such as stream fusion~\cite{coutts2007stream} use local rewrite rules to remove intermediate arrays, and tend to rely on general purpose compiler optimisations to expose opportunities for these rewrites to occur.
Generally, the local rewrite rule can only be applied once the producer has been inlined into the consumer.
Because of the local nature of these rewrites and the reliance on inlining, local fusion can only perform producer-consumer fusion with a single consumer, as inlining the producer into multiple consumers would duplicate work.
Local fusion is fragile because heuristics are used to determine whether inlining occurs~\cite{lippmeier2013data}.
This fragility poses serious problems to programmers who require fusion for adequate performance, as one must have deep knowledge of the internal compiler phases and library design in order to predict whether fusion will occur -- or worse yet, they must scour the intermediate code to count the number of loops produced. 
In contrast, global fusion systems such as those traditionally used in imperative compilers make use of specific optimisations which must be implemented separately, but are able to perform horizontal fusion and are less prone to the fragility of local fusion.

This local/global distinction is not a hard classification, as some systems fit somewhere in between.
For example, {\bf strymonas}~\cite{kiselyov2016stream} uses staged compilation for introspection of the code and to ensure appropriate inlining, but only performs producer-consumer fusion with a single local rule.
While not explicitly mentioned in the paper, Kiselyov et al~\cite{kiselyov2016stream}'s {\bf strymonas} system only allows streams to be used once.
Furthermore, the entire stream computation must be terminated with a single fold which takes a single stream.
This means multiple outputs with different rates cannot be treated, for example @partition@ which returns multiple arrays filtered with different predicates.

This innocent-looking yet useless program, which should pair each line with itself, actually pairs each line with the successive line.
\begin{code}
let lines = linesFromFile filename
in  zip lines lines
\end{code}

One would expect the above program to have the same behaviour as the following - or at least produce an error otherwise:
\begin{code}
zip (linesFromFile filename) (linesFromFile filename)
\end{code}

The one feature they support that we do not is concatMap/flatMap, which concatenates all produced substreams.
This is a limitation in the \emph{conversion} to process calculus, not in the process calculus itself.
Our process calculus can handle concrete instantiations of concatMap when specialised to a particular subprocess, but not the general case where the subprocess is statically unknown.
By using a staging restriction as they do, we could modify the conversion to create a specialised version of concatMap for the given argument.

\subsection{The problem with synchronised product}
\label{s:Synchro}
Related work on Network Fusion~\cite{fradet2004network} allows fusion of Kahn Process Networks using synchronised product to fuse pairs of processes together.
Synchronised product is: both machines can take independent steps so long as it is not in the alphabet of the other, but have to agree on any shared actions.
So if both machines pull from @xs@, the @pull xs@ message can only be executed when both machines agree.

This sort of coordination is a bit too coarse and causes deadlocks.
This example cannot be fused by synchronised product: one zip is trying to pull from @as@ and the other from @bs@, but neither can proceed without the other.
Note that while this contrived example is unlikely to be written by hand, it is plausible that it would be part of a larger program after inlining has occurred.

\begin{code}
zips :: [a] -> [b] -> [a*b] * [b*a]
zips as bs =
  let abs = zip as bs
      bas = zip bs as
  in  abs, bas
\end{code}

When stream programs are encoded as labelled transition systems, some fusion can be done by computing the synchronised product of two transition systems, but this suffers deadlock when the two programs share multiple inputs and read them in different orders.
By annotating transition states with the status of each input channel, our fusion algorithm works for these cases where synchronised product does not.

% After inlining the definition of zip, we have:
% \begin{code}
% zips :: [a] -> [b] -> [a*b] * [b*a]
% zips as bs =
%   let abs = process abs.
%        let loop1 =
%           a <- pull as
%           b <- pull bs
%           push abs (a,b)
%           drop as
%           drop bs
%           jump loop1
%        in jump loop1
%       bas = process bas.
%        let loop2 =
%           b <- pull bs
%           a <- pull as
%           push abs (b,a)
%           drop bs
%           drop as
%           jump loop2
%        in jump loop2
%   in  abs, bas
% \end{code}
% 
% Here, in order to compute the synchronised product of these two, we need to know which are the common or shared actions.
% Shared are: @pull as@, @pull bs@, @drop as@ and @drop bs@.
% So both processes have to execute these together: if one wants to @pull as@, it can only act when the other one wants to @pull as@ as well.
% Thus, the two machines are deadlocked, waiting for the other one to agree.
% 
% For this case, a simple solution suffices to fix this: adding extra processes for splits. So we would add a new combinator @dup@ which takes one input stream and creates two output streams. It reads a single element from the input and pushes it to both outputs.
% 
% \begin{code}
% dup = stream_1_2 \is ols ors.
%   letrec
%     p1   = pull is p2
%     p2 i = push ols i (p3 i)
%     p3 i = push ors i p4
%     p4   = drop is p1
%   in p1
% \end{code}
% And the modified zip uses this like so:
% 
% \begin{code}
% zips :: [a] -> [b] -> [a*b] * [b*a]
% zips as bs =
%   let as1, as2 = dup as
%       bs1, bs2 = dup bs
%       abs = zip as1 bs1
%       bas = zip bs2 as2
%   in  abs, bas
% \end{code}
% 
% This has removed the deadlock, allowing a trace such as:
% \begin{code}
% as, as1, bs, bs1, bs2, as2
% \end{code}
% 
% However it is not hard to modify the program so that the deadlock still exists.
% By changing the each zip to pull from @as2@ and @bs2@ first, the streams are still used linearly, but deadlock occurs as @as1@ cannot be produced until after @bs2@ has been consumed, and vice versa.
% \begin{code}
% zips :: [a] -> [b] -> [a*b] * [b*a]
% zips as bs =
%   let as1, as2 = dup as
%       bs1, bs2 = dup bs
%       abs = zip as2 bs1
%       bas = zip bs2 as1
%   in  abs, bas
% \end{code}
% 
% Using @dup@ we can concoct a simpler example with two processes and one input stream where synchronised product deadlocks.
% \begin{code}
% zipself :: [a] -> [a*a]
% zipself as =
%   let as1, as2 = dup as
%       aas = zip as2 as1
%   in  aas
% \end{code}
% Here @dup@ is attempting to push to @as1@, but cannot until @zip@ pulls, while @zip@ is attempting to pull from @as2@, which cannot proceed until the @dup@ pushes.

>>>>

This work aims to address the problems with existing combinator-based array fusion systems.
As stated in the introduction, neither pull-based or push-based fusion alone is sufficient.
Some combinators are inherently push-based, particularly those with multiple outputs such as @unzip@; while others are inherently pull-based, such as @zip@.

Short cut fusion is an attractive idea, as it allows fusion systems to be specified with a single rewrite rule.
However, short cut fusion relies on inlining which, like pull-based streams, only occurs when there is a single consumer.
Thus, short cut fusion is inherently biased towards pull fusion.
Push-based short cut fusion systems \emph{do} exist \cite{gill1993short}, but support only the intersection of pull and push: that is, neither @zip@ nor @unzip@ \cite{svenningsson2002shortcut,lippmeier2013data}.

Recent work on stream fusion by \citet{kiselyov2016stream} uses staged computation in a push-based system to ensure all combinators are inlined, but when streams are used multiple times this causes excessive inlining, which duplicates work.
For effectful input streams, for example reading from the network, this duplicated work can change the overall meaning of the program.
% I could write more about this eg only supporting a single output, but the other points probably apply to push streams in general

Data flow fusion~\cite{lippmeier2013data} is neither pull-based nor push-based, and supports arbitrary splits and joins.
It supports standard combinators such as @map@, @filter@ and @fold@, and converts each stream to a series with explicit rate types, similar to the clock types of Lucid Synchrone \cite{benveniste2003synchronous}.
These rate types ensure that well-typed programs can be fused without introducing unbounded buffers.
This allows unfusable programs to be caught at compile time.
The main drawback of data flow fusion is that it only supports a limited set of combinators, and adding more is non-trivial.

One way to address the difference between pull and push streams is to explicitly support both separately, as seen in \citet{bernardy2015duality} and \citet{lippmeier2016polarized}.
Here, pull streams have the type @Source@ and are represented as a source that is always available to be pulled from, while push streams have the type @Sink@ and are represented as a sink that can always accept input.
Both systems rely on stream bindings being used linearly to ensure correctness, including boundedness of buffers and ordering of effects.
Operations over sources are expressed fairly naturally compared to streams, for example the @zip@ combinator has the type @Source a -> Source b -> Source (a,b)@.
Sinks, however, are co-variant, and operations must be performed somewhat backwards, so that the @unzip@ combinator takes the two output sinks to push into and returns a new sink that pushes into these.
It has the type @Sink a -> Sink b -> Sink (a,b)@.
This system requires the streaming computation to be manually split into sources and sinks, and be joined together by a loop that `drains' values from the source and pushes them into the sink.

The duality between pull and push arrays has also been explored in Obsidian by \citep{claessen2012expressive} and later in \citep{svensson2014defunctionalizing}.
Here the distinction is made for the purpose of code generation for GPUs rather than fusion, as operations such as appending pull arrays require conditionals inside the loop, whereas using push arrays moves these conditionals outside the loop.

Streaming IO libraries have blossomed in the Haskell ecosystem, generally based on Iteratees \cite{kiselyov2012iteratees}.
Libraries such as @conduit@ \cite{hackage:conduit}, @enumerator@ \cite{hackage:enumerator}, @machines@ \cite{hackage:machines} and @pipes@ \cite{hackage:pipes} are all designed to write stream computations with bounded buffers.
However, these libraries provide no fusion guarantees, and as such programs tend to be written over chunks of data to make up for the communication overhead.
For the most part they support only straight-line computations, with only limited forms of branching.

In relation to process calculi, synchronised product has been suggested as a method for fusing Kahn process networks together~\cite{fradet2004network}, but does not appear to have been implemented or evaluated.
The synchronised product of two processes allows either process to take independent or local steps at any time, but shared actions, such as when both processes communicate on the same channel, must be taken in both processes at the same time.
This is a much simpler fusion method than ours, but is also much stricter.
When two processes share multiple channels, synchronised product will fail unless both processes read the channels in exactly the same order.
Our system can be seen as an extension of synchronised product that allows some leeway in when processes must take shared steps: they do not have to take shared steps at the same time, but if one process lags behind the other, it must catch up before the other one gets too far ahead.

Synchronous languages such as LUSTRE~\cite{halbwachs1991synchronous}, Lucy-n~\cite{mandel2010lucy} and SIGNAL~\cite{le2003polychrony} all use some form of clock calculus and causality analysis to ensure that programs can be statically scheduled with bounded buffers.
These languages describe \emph{passive} processes where values are fed in to streams from outside environments, such as data coming from sensors.
In this case, the passive process has no control over the rate of input coming in, and if they support multiple input streams, they must accept values from them in any order.
In contrast, the processes we describe are \emph{active} processes that have control over the input that is coming in.
This is necessary for combinators such as mergesort-style @merge@, as well as @append@.
Note that in the synchronous language literature, it is common to refer to a different merge operation, also known as @default@, which computes a stream that is defined whenever either input is defined.

Synchronous dataflow (not to be confused with synchronous languages above) is a dataflow graph model of computation where each dataflow actor has constant, statically known input and output rates.
The main advantage of synchronous dataflow is that it is simple enough for static scheduling to be decidable, but this comes at a cost of expressivity.
StreamIt~\cite{thies2002streamit} uses synchronous dataflow for scheduling when possible, otherwise falling back to dynamic scheduling~\cite{soule2013dynamic}.
Boolean dataflow and integer dataflow~\cite{buck1993scheduling,buck1994static} extend synchronous dataflow with boolean and integer valued control ports, and attempt to recover the structure of ifs and loops from select and switch actors.
These systems allow some dynamic structures to be scheduled statically, but are very rigid and only support limited control flow structures: it is unclear how merge or append could be scheduled by this system.
Finite state machine-based scenario aware dataflow (FSM-SADF)~\cite{stuijk2011scenario,van2015scenario} is still quite expressive compared to boolean and integer dataflow, while still ensuring static scheduling.
A finite state machine is constructed, where each node of the FSM denotes its own synchronous dataflow graph.
The FSM transitions from one dataflow graph to another based on control outputs of the currently executing dataflow graph.
For example, a filter is represented with two nodes in the FSM.
The dataflow graph for the initial state executes the predicate, and the value of the predicate is used to determine which transition the FSM takes: either the predicate is false and the FSM stays where it is, or the predicate is true and moves to the next state.
The dataflow graph for the next state emits the value, and moves back to the first state.
This does appear to be able to express value-dependent operations such as merge, but lacks the composability - and familiarity - of combinators.

% StreamIt:
% Only allows limited splits and joins: round robin and duplication for splits, round robin and combination for joins. 
% Does not support fully general graphs - instead using combinators to introduce a (split/join) and a combinator for a feedback loop.
% 
% Parameterized dataflow (PDF),  \cite{bhattacharya2001parameterized}
% Schedulable parametric dataflow (SPDF),  \cite{fradet2012spdf}
